An Extension of LTL with Rules\\ and its Application to Runtime Verification

### 2019-07-15 conclusion.tex
Updates from Overleaf

@@

Propositional linear temporal logic (LTL) and automata are two common specification formalisms for software and hardware systems. While temporal logic has a more declarative
flavor, automata are more operational, describing how the specified system
progresses. There has been several proposed extensions to LTL that extend
its expressive power to that of related automata formalisms. We proposed here
a simple extension for propositional LTL that is based on adding propositions that summarize the prefix of an execution. Conceptually, this extension puts the specification in between propositional LTL and automata, as the additional
variables can be seen as representing the state of an automaton
that is synchronized with the temporal property. It is shown to have
the same expressive power as automata models, and is in particular
appealing for runtime verification of past (i.e., safety) temporal properties,
which already are based on summarizing the value of subformulas over
observed prefixes.
%
We demonstrated that first-order linear temporal logic (\FLTL{}), which
can be used to assert properties about systems with data, also has expressiveness deficiencies, and extended it with relations that summarize prefixes.
We showed that for the first-order case, unlike the propositional case,
this extension is not identical to the addition of dynamic (i.e., state dependent) quantification.
%
We presented a monitoring
algorithm for propositional past time temporal logic, extending a classical algorithm, and similarly presented an algorithm for 
first-order past temporal logic.
Finally we described the implementation of this extension in the \dejavu{} tool and provided experimental results.
%
Future work includes performing additional experiments, and making further comparisons to other formalisms. Finally we would like to study further extensions, exploring the space between logic and programming.

>>>>

\vspace{-1.7cm} \quad

@@

Propositional linear temporal logic (LTL) and automata are two common specification formalisms for software and hardware systems. While temporal logic has a more declarative
flavor, automata are more operational, describing how the specified system
progresses. There has been several proposed extensions to LTL that extend
its expressive power to that of related automata formalisms.

We proposed here
a simple extension for propositional LTL that adds auxiliary propositions that summarize the prefix of the execution, based on rules written as past formulas. Conceptually, this extension puts the specification in between propositional LTL and automata, as the additional
variables can be seen as representing the state of an automaton
that is synchronized with the temporal property. It is shown to have
the same expressive power as B\"{u}chi automata, and is in particular
appealing for runtime verification of past (i.e., safety) temporal properties,
which already are based on summarizing the value of subformulas over
observed prefixes.

%
We demonstrated that first-order linear temporal logic (\FLTL{}), which
can be used to assert properties about systems with data, also has expressiveness deficiencies, and similarly
extended it with rules
that define {\em relations} that summarize prefixes of the execution.
We proved that for the first-order case, unlike the propositional case,
this extension is not identical to the addition of dynamic (i.e., state dependent) quantification.
%
We presented a monitoring
algorithm for propositional past-time temporal logic with rules, extending a classical algorithm, and similarly presented an algorithm for 
first-order past temporal logic with rules.
Finally we described the implementation of this extension in the \dejavu{} tool and provided experimental results. The
code and many more examples appear on:
\begin{center}
\url{https://github.com/havelund/dejavu}
\end{center}

%
Future work includes performing additional experiments, and making further comparisons to other formalisms. We intend to study further extensions, exploring the space between logic and programming.

### 2019-07-15 implementation.tex
Updates from Overleaf

\vspace{1ex}\noindent {\bf Example properties.}
Figure \ref{fig:properties} shows three properties in the 
input ASCII format of the tool, related to the examples in Section~\ref{sec:syntax-semantics}, which
are not expressible in (P)FLTL.
These properties are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}. 
%
The ASCII version of the logic uses 
\idsl{@} for $\ominus$,
\idsl{|} for $\vee$,
\idsl{&} for $\wedge$, and
\idsl{!} for $\neg$.
%
The first property \idsl{telemetry1} is a variant of
formula \ref{eq:wolper-first-order}, illustrating the use of a 
rule to express a first-order version of Wolper's example \cite{Wolper}, that all the states with even indexes of a sequence satisfy a property.
In this case we consider a radio on board a spacecraft, which communicates over different channels (quantified over in the formula), which can be turned on and off with a \idsl{toggle(x)} - they are initially off.
Telemetry can only be sent to ground over a channel \idsl{x} with the \idsl{telem(x)} event when radio channel \idsl{x} is toggled on.

%%%%

The second property, \idsl{telemetry2}, expresses the same 
property as \idsl{telemetry1}, but in this case using two 
rules, more closely reflecting how we would model this using a state machine with two states for each channel \idsl{x}: 
\idsl{closed(x)} and \idsl{open(x)}. The rule \idsl{closed(x)} is defined as a disjunction between three alternatives. The first states that this predicate is true if we are in the initial state 
(the only state where \idsl{@true} is false), and there is no \idsl{toggle(x)} event. The next alternative states that
\idsl{closed(x)} was true in the previous state and there is
no \idsl{toggle(x)} event.
The third alternative states that in the previous state we were
in the \idsl{open(x)} state and we observe a \idsl{toggle(x)} 
event. Similarly for the \idsl{open(x)} rule.

@@

prop spawning :
  Forall x . Forall y . Forall d . report(y,x,d) -> spawned(x,y) where
    spawned(x,y) := @ spawned(x,y) | spawn(x,y) | 
                    Exists z . (@spawned(x,z) & spawn(z,y))
\end{lstlisting}
\caption{Three properties stated in \dejavu's logic}
\label{fig:properties}
\end{center}
\end{figure}

@@

\vspace{1ex}\noindent {\bf Evaluation.}
We evaluated the three properties in Figure 
\ref{fig:properties} on a collection of traces. Table \ref{tab:evaluation} shows the  
analysis time (excluding time to compile the generated 
monitor) for different traces (format is `trace length : seconds'). The telemetry traces will repeatedly open hundreds of channels and send telemetry over them. 
%
The spawning traces consist of 50\% spawning events and 50\% reporting events. The spawning property  requires larger processing time compared to the telemetry properties since more data (the transitive closure) has to be computed and stored.

@@

\ignore{
  // threads: how many threads initially spawned by main program
  // repeat: how many times new threads from new threads are spawned
  // number of spawned threads correspond to number of events
}

\begin{table}[h!]
\centering
\begin{tabular}{|l|c|c|c|}
\hline
Property   & Trace 1 & Trace 2 & Trace 3 \\
\hline
telemetry1 & 1,200,001\,:\,3.3s & 5,200,001\,:\,7.1s & 10,200,001\,:\,12.8s \\
\hline
telemetry2 & 1,200,001\,:\,4.4s & 5,200,001\,:\,9.9s & 10,200,001\,:\,17.9s \\
\hline
spawning   & 9,899\,:\,42.7s & 19,999\,:\,127.6s & 39,799\,:\,572.4s \\
\hline
\end{tabular}
\caption{Evaluation - trace lengths and analysis time in seconds}
\label{tab:evaluation}
\end{table}

>>>>

@@

prop spawning : Forall x . Forall y . Forall d . report(y,x,d) -> spawned(x,y) where
  spawned(x,y) := @spawned(x,y) | spawn(x,y) | 
                  Exists z . (@spawned(x,z) & spawn(z,y))

prop commands : Forall c . dispatch(c) -> ! already_dispatched(c) where
  already_dispatched(c) := @ [dispatch(c) , complete(c)),
  dispatch(c) := Exists t . CMD_DISPATCH(c,t),
  complete(c) := Exists t . CMD_COMPLETE(c,t)
\end{lstlisting}
\caption{Properties stated in \dejavu's logic}
\label{fig:properties}
\end{center}
\end{figure}

\vspace{1ex}\noindent {\bf Example properties.}
Figure \ref{fig:properties} shows four properties in the 
input ASCII format of the tool, the first three of which are related 
to the examples in Section~\ref{sec:syntax-semantics}, which
are not expressible in (P)\FLTL{}. That is,
these properties are not expressible in the original first-order 
logic of \dejavu{} presented in \cite{HPU}. The last property illustrates the use of rules to perform conceptual abstraction.
%
The ASCII version of the logic uses 
\idsl{@} for $\ominus$,
\idsl{|} for $\vee$,
\idsl{&} for $\wedge$, and
\idsl{!} for $\neg$.
%
The first property \idsl{telemetry1} is a variant of
formula \ref{eq:wolper-first-order}, illustrating the use of a 
rule to express a first-order version of Wolper's example \cite{Wolper}, that all the states with even indexes of a sequence satisfy a property.
In this case we consider a radio on board a spacecraft, which communicates over different channels (quantified over in the formula) that can be turned on and off with a \idsl{toggle(x)}; they are initially off.
Telemetry can only be sent to ground over a channel \idsl{x} with the \idsl{telem(x)} event when radio channel \idsl{x} is toggled on.

The second property, \idsl{telemetry2}, expresses the same 
property as \idsl{telemetry1}, but in this case using two 
rules, reflecting how we would model this using a state machine with two states for each channel \idsl{x}: 
\idsl{closed(x)} and \idsl{open(x)}. The rule \idsl{closed(x)} is defined as a disjunction between three alternatives. The first states that this predicate is true if we are in the initial state 
(the only state where \idsl{@true} is false), and there is no \idsl{toggle(x)} event. The next alternative states that
\idsl{closed(x)} was true in the previous state and there is
no \idsl{toggle(x)} event.
The third alternative states that in the previous state we were
in the \idsl{open(x)} state and we observe a \idsl{toggle(x)} 
event. Similarly for the \idsl{open(x)} rule.

The fourth property, \idsl{commands}, 
concerns a realistic log from from the Mars rover Curiosity 
\cite{msl}. The log consists of events 
(here renamed) \idsl{CMD_DISPATCH(c,t)} and \idsl{CMD_COMPLETE(c,t)}, 
representing the dispatch and subsequent completion  of a command 
\idsl{c} at time \idsl{t}. The property to be verified is that a command, once 
dispatched, is not dispatched again before completed. Rules 
are used to break down the formula to conceptually simpler pieces. 
%The example
%illustrates how rules elegantly replace the role of macros, a non-trivial extension 
%introduced in an earlier version of \dejavu.

% Out of 25,463 dispatches and 23,740 completions, there were 2,466
% violations of this property.
%
% A formulation of this property without using rules is:
%
% prop commands :
%   Forall c .
%     (Exists t1 . CMD_DISPATCH(c,t1)) -> 
%       ! @ [
%             Exists t2 . CMD_DISPATCH(c,t2), 
%             Exists t3 . CMD_COMPLETE(c,t3)
%           )

@@

\vspace{1.5ex}\noindent {\bf Evaluation.}
In \cite{HPU,HPU-FMSD} we evaluated \dejavu{}
without the rule extension against the \monpoly{} tool \cite{agrebasin}, which supports a logic close to \dejavu's. In \cite{HP} we evaluated \dejavu's garbage collection capability. In this section we evaluate the rule extension for the properties in Figure 
\ref{fig:properties} on a collection of 
traces. Table \ref{tab:evaluation} shows the  
analysis time (excluding time to compile the 
generated monitor) and maximal memory usage in 
MB for different 
% ---
traces
%\footnote{
%For the first two properties the same traces were used, generated as follows for varying constants $P$, $Q$ and $R$.
%Repeat $P$ times the following: toggle $Q$ 
%channels; then for each channel send telemetry $R$ %times; then toggle the channels again.
%For the third property, traces were 
%generated as follows for varying constants $S$ 
%and $T$: spawn $S$ threads from the main thread 0, and %let each report back to thread 0; then repeat $T$ times %the following: each newly spawned thread in previous %step
%spawns a new thread, which reports back to the main %thread 0.
%}
% ---
(format is `trace length : time / memory'). 
%
%The telemetry traces will repeatedly open 
%hundreds of channels and send telemetry over 
%them. 
%
%The spawning traces consist of 50\% spawning 
%events and 50\% reporting events. 
%
The processing time is generally
very reasonable for very large traces. However,
the spawning property  requires considerably larger 
processing time and memory compared to the other 
properties since more data (the transitive 
closure) has to be computed and stored.
%
The evaluation was performed on a Mac laptop, with the Mac OS X 10.10.5 operating 
system, on a 2.8 GHz Intel Core i7 with 16 GB of memory.

@@

\ignore{
  // threads: how many threads initially spawned by main program
  // repeat: how many times new threads from new threads are spawned
  // number of spawned threads correspond to number of events
}
\begin{table}[h!]
\centering
\begin{tabular}{|l|c|c|c|}
\hline
Property   & Trace 1 & Trace 2 & Trace 3 \\
\hline
telemetry1 & 1,200,001\;:\;2.6s / 194 MB & 5,200,001\;:\;5.9s / 210 MB & 10,200,001\;:\;10.7s / 239 MB \\
\hline
telemetry2 & 1,200,001\;:\;3.8s / 225 MB & 5,200,001\;:\;8.7s / 218 MB & 10,200,001\;:\;16.6s / 214 MB \\
\hline
spawning   & 9,899\;:\;29.5s / 737 MB & 19,999\;:\;117.3s / 1,153 MB & 39,799\;:\;512.5s / 3,513 MB \\
\hline
commands   & 49,999\;:\;1.5s / 169 MB & N/A & N/A \\
\hline
\end{tabular}
\caption{  Evaluation - trace lengths, analysis time in seconds, and maximal memory use}
\label{tab:evaluation}
\end{table}

%\begin{table}[h!]
%\centering
%\begin{tabular}{|l|c|c|c|}
%\hline
%Property   & Trace 1 & Trace 2 & Trace 3 \\
%\hline
%telemetry1 & 1,200,001\,:\,3.3s & 5,200,001\,:\,7.1s & 10,200,001\,:\,12.8s \\
%\hline
%telemetry2 & 1,200,001\,:\,4.4s & 5,200,001\,:\,9.9s & 10,200,001\,:\,17.9s \\
%\hline
%spawning   & 9,899\,:\,42.7s & 19,999\,:\,127.6s & 39,799\,:\,572.4s \\
%\hline
%\end{tabular}
%\caption{Evaluation - trace lengths and analysis time in seconds}
%\label{tab:evaluation}
%\end{table}

### 2019-07-15 main.tex
Updates from Overleaf

\documentclass{llncs}
%\documentclass[converence,10pt]{IEEEtran}
%\IEEEoverridecommandlockouts
%\usepackage{cite}
%\usepackage{amsmath,amssymb,amsfonts}
%\usepackage{algorithmic}
%\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{bm}
%\usepackage{xcolor}
%\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
%    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\newcommand{\Exists}{\bm{\exists}\kern-0.6em\bm{\exists}}
\newcommand{\Forall}{\bm{\forall}\kern-0.6em\bm{\forall}}

@@

\ignore{\author{
\begin{tabular}{ccc}
Klaus Havelund & \vspace{0.2cm} & 
Doron Peled \\
NASA Jet Propulsion Laboratory, Pasadena, USA & & 
Bar Ilan University, Ramat Gan, Israel \\
klaus.havelund@jpl.nasa.gov & & doron.peled@gmail.com
\end{tabular}
}
}

\ignore{
\author{\IEEEauthorblockN{Klaus Havelund} 
\IEEEauthorblockA{\textit{Laboratory for Reliable 
Software} \\
\textit{NASA Jet Propulsion Laboratory}\\
Pasadena, USA \\
klaus.havelund@jpl.nasa.gov}\\\mbox{}\\
\and
\IEEEauthorblockN{Doron Peled} 
\IEEEauthorblockA{\textit{Dep. of Computer Science} \\
\textit{Bar Ilan University}\\
Ramat Gan, Israel \\
doron.peled@gmail.com}
}
}

@@

\iffalse
\author{ 
Klaus Havelund~\thanks{The research performed by this author was carried out at Jet Propulsion Laboratory, California Institute of Technology, under a contract with the National Aeronautics and Space Administration.} 
\and
Doron Peled~\thanks{The research performed by this author was partially funded by Israeli Science Foundation grant 1464/18: 
``Efficient Runtime Verification for Systems with Lots of Data and its Applications''.
}
}

\institute{
Jet Propulsion Laboratory, \\
California Institute of Technology, USA 
\and
Department of Computer Science \\
Bar Ilan University, Israel
}

\fi

@@

\begin{abstract}

Runtime Verification (RV) consists of analyzing execution 
traces using formal techniques. 
%RV is scalable and 
%therefore has a big potential for infusion in software 
%engineering practices. 
It includes monitoring the 
execution of a system against properties formulated in 
Linear Temporal Logic (LTL). 
LTL offers a succinct notation for writing useful 
specification properties. However, it is limited in 
expressiveness in the propositional case, and several 
theoretic extensions have therefore been proposed. Furthermore, for many practical cases, there is a need to monitor properties 
that carry data, where one can use formalisms like 
first-order LTL. We show that 
first-order LTL has similar expressiveness
limitations as the propositional version. We suggest here 
two related extensions for increasing the expressive power: 
one for propositional LTL and one for 
first-order LTL. These extensions have a simple incremental operational 
semantics that is more suitable for RV than previously 
suggested extensions. We show that the 
propositional extension has the same expressiveness as the 
classical extensions, and demonstrate its
adoption for runtime verification of first-order safety 
properties. Finally, we expand the BDD-based runtime verification 
tool \dejavu{}\footnote{The actual tool name is not given due to anonymous reviewing.}  to support our extension and perform some 
experiments.

\end{abstract}

@@

@@

%%%M
\subsection*{Details of left to right direction of Theorem~\ref{theo1}.}
One can translate any \QLTL{} property into a second-order monadic logic in a standard
way~\cite{Thomas}, and then from
second-order monadic logic to a deterministic Muller automaton $M = (S, \iota , \Sigma, \Delta, {\cal F})$ (a B\"{u}chi automaton may not always be deterministic), where $S$ is the set of states, $\iota \in S$ is the initial state, $\Sigma$ is the set of inputs, each $\alpha \in \Sigma$ is a minterm (i.e., conjunction of
each variable in the set negated or non-negated) of the variables in $B$, $\Delta : S \times \Sigma \mapsto S$ is the transition function, and ${\cal F} = \{ F_1 , \ldots , F_k \}$ is the accepting set. A muller automaton 
accepts a run if it 
passes infinitely often in exactly
one of the specified accepting sets. To formulate an equivalent \ELTL{} formula,
the auxiliary variables represent the states $S$ of $M$, i.e., $a_s$ represents the state $s \in S$.
Then we translate $M$ into an \ELTL{} formula
of the form:

@@

\begin{itemize}
\item $I [ \varphi , {\cal R} , 0 ] = \emptyset$.
\item $I [ \true , {\cal R} , i ] = \{ \epsilon \}$.
\item $I [ p ( a ) , {\cal R} , i ] =$ if $m (p )  ( a , i ) $ then
$\{ \epsilon \}$ else $\emptyset$.
\item $I [ p ( x ) , {\cal R} , i ] = \{ [ x \mapsto a ] | m (p) ( a , i)  \}$.
%\item $I [ \true , \sigma , i ] = D^n$.
%\item $I [ ( \varphi \vee \psi ) , \sigma , i ] = 
%I [ \varphi , \sigma , i ] \cup I [ \psi , \sigma , i ]$.
\item $I [ ( \varphi \wedge \psi ) , \sigma , i ] = 
I [ \varphi , {\cal R} , i ] \;  \bigcap \; I [ \psi , \sigma , i ]$.
\item $I [ \neg \varphi , {\cal R} , i ] = 
A_{\vars ( \varphi )} \; \setminus \; I [ \varphi , {\cal R} , i ]$.
\item $I [ ( \varphi \; \since \; \psi ) , {\cal R} , i ] = 
I [ \psi , {\cal R}, i ] \; \bigcup \;
( I [ \varphi , {\cal R} , i ] \bigcap 
I [ (\varphi \since \psi ) , {\cal R} , i - 1 ] )$.
\item $I [ \ominus \varphi , {\cal R} , i ] = I [ \varphi , {\cal R} , i-1 ]$.
\item $I [ \exists x \; \varphi , \sigma , i ] = 
\proj ( I [ \varphi , {\cal R} , i ] , \{ x \} )$.
\item $I [ \psi \mathrm{\ where\ } r_j ( x_j) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} , {\cal R} , i ] = 
 I [ \psi , {\cal Q} ] $ where ${\cal Q} = {\cal R} \cup
\{ r_j | r_j [ i ] =  I [ 
\varphi_j , {\cal Q} , i ] \}$
\end{itemize}
Note that the last time interprets a formula of
the form defined in~(\ref{EFLTL}).
%\klaus{Above: as the semantics is presented, it looks as if 
%the `$\psi\ where ...$' is just another subformula.}
%\klaus{Above: hide is not defined.}
%\klaus{Above: the semantics of $p(a)$ and $p(x)$ do not refer to 
%${\cal R} $.}
In the last item, the extended model ${\cal Q}$ appears both on the left and the right of the equality sign. It is well defined because of Lemma~\ref{samesame} (equivalently, a more verbose definition can include a stepwise construction of ${\cal Q}$ from ${\cal R}$).
\end{document}

>>>>

\documentclass{llncs}
\usepackage{authblk}
%\documentclass[converence,10pt]{IEEEtran}
%\IEEEoverridecommandlockouts
%\usepackage{cite}
%\usepackage{amsmath,amssymb,amsfonts}
%\usepackage{algorithmic}
%\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{bm}
%\usepackage{xcolor}
%\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
%    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\newcommand{\Exists}{\bm{\exists}\kern-0.6em\bm{\exists}}
\newcommand{\Forall}{\bm{\forall}\kern-0.6em\bm{\forall}}

@@

%\ignore{
%\author{\IEEEauthorblockN{Klaus Havelund} 
%\IEEEauthorblockA{\textit{Laboratory for Reliable 
%Software} \\
%\textit{NASA Jet Propulsion Laboratory}\\
%Pasadena, USA \\
%klaus.havelund@jpl.nasa.gov}\\\mbox{}\\
%\and
%\IEEEauthorblockN{Doron Peled} 
%\IEEEauthorblockA{\textit{Dep. of Computer Science} \\
%\textit{Bar Ilan University}\\
%Ramat Gan, Israel \\
%doron.peled@gmail.com}
%}

@@

\author{Klaus Havelund$^1$ \and Doron Peled$^2$}

%\ignore{
%Klaus Havelund~\thanks{The research performed by this %author was carried out at Jet Propulsion Laboratory, %California Institute of Technology, under a contract %with the National Aeronautics and Space %Administration.} 
%\and
%Doron Peled
%~\thanks{The research performed by this author was %partially funded by Israeli Science Foundation grant %1464/18: 
%``Efficient Runtime Verification for Systems with Lots %of Data and its Applications''.

\institute{
Jet Propulsion Laboratory, 
California Institute of Technology, USA 
\and
Department of Computer Science,
Bar Ilan University, Israel
}

\iffalse
\author{
\begin{tabular}{ccc}
Klaus Havelund & \vspace{0.2cm} & 
Doron Peled \\
NASA Jet Propulsion Laboratory, Pasadena, USA & & 
Bar Ilan University, Ramat Gan, Israel \\
klaus.havelund@jpl.nasa.gov & & doron.peled@gmail.com
\end{tabular}
}
\fi

@@

\input{abstract}

\input{introduction}

@@

%\newpage

@@

%%%M
\subsection*{Details of left to right direction of Theorem~\ref{theo1}.}
One can translate any \QLTL{} property into a second-order monadic logic in a standard
way~\cite{Thomas}, and then from
second-order monadic logic to a deterministic Muller automaton $M = (S, \iota , \Sigma, \Delta, {\cal F})$ (a deterministic B\"{u}chi automaton may not always be possible), where $S$ is the set of states, $\iota \in S$ is the initial state, $\Sigma$ is the set of inputs, each $\alpha \in \Sigma$ is a minterm (i.e., conjunction of
each variable in the set negated or non-negated) of the variables in $B$, $\Delta : S \times \Sigma \mapsto S$ is the transition function, and ${\cal F} = \{ F_1 , \ldots , F_k \}$ is the accepting set. A muller automaton 
accepts a run if it 
passes infinitely often in exactly
one of the specified accepting sets. To formulate an equivalent \ELTL{} formula,
the auxiliary variables represent the states $S$ of $M$, i.e., $a_s$ represents the state $s \in S$.
Then we translate $M$ into an \ELTL{} formula
of the form:

@@

\begin{itemize}
\item $I [ \varphi , {\cal R} , 0 ] = \emptyset$.
\item $I [ \true , {\cal R} , i ] = \{ \epsilon \}$.
\item $I [ p ( a ) , {\cal R} , i ] =$ if $m (p )  ( a , i ) $ then
$\{ \epsilon \}$ else $\emptyset$.
\item $I [ p ( x ) , {\cal R} , i ] = \{ [ x \mapsto a ] | m (p) ( a , i)  \}$.
%\item $I [ \true , \sigma , i ] = D^n$.
%\item $I [ ( \varphi \vee \psi ) , \sigma , i ] = 
%I [ \varphi , \sigma , i ] \cup I [ \psi , \sigma , i ]$.
\item $I [ ( \varphi \wedge \psi ) , \sigma , i ] = 
I [ \varphi , {\cal R} , i ] \;  \bigcap \; I [ \psi , \sigma , i ]$.
\item $I [ \neg \varphi , {\cal R} , i ] = 
A_{\vars ( \varphi )} \; \setminus \; I [ \varphi , {\cal R} , i ]$.
\item $I [ ( \varphi \; \since \; \psi ) , {\cal R} , i ] = 
I [ \psi , {\cal R}, i ] \; \bigcup \;
( I [ \varphi , {\cal R} , i ] \bigcap 
I [ (\varphi \since \psi ) , {\cal R} , i - 1 ] )$.
\item $I [ \ominus \varphi , {\cal R} , i ] = I [ \varphi , {\cal R} , i-1 ]$.
\item $I [ \exists x \; \varphi , \sigma , i ] = 
\proj ( I [ \varphi , {\cal R} , i ] , \{ x \} )$.
\item $I [ \psi \mathrm{\ where\ } r_j ( x_j) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} , {\cal R} , i ] = 
 I [ \psi , {\cal Q} ] $ where ${\cal Q} = {\cal R} \cup
\{ r_j | r_j [ i ] =  I [ 
\varphi_j , {\cal Q} , i ] \}$
\end{itemize}
Note that the last time interprets a formula of
the form defined in~(\ref{EFLTL}).
%\klaus{Above: as the semantics is presented, it looks as if 
%the `$\psi\ where ...$' is just another subformula.}
%\klaus{Above: hide is not defined.}
%\klaus{Above: the semantics of $p(a)$ and $p(x)$ do not refer to 
%${\cal R} $.}
In the last item, the extended model ${\cal Q}$ appears both on the left and the right of the equality sign. It is well defined because of Lemma~\ref{samesame} (equivalently, a more verbose definition can include a prefix-inductive construction of ${\cal Q}$ from ${\cal R}$).
\end{document}
